Joint analysis of read count data

### 2020-09-11 combined_analysis/combine_all_DLPFC.Rmd
Inital commit

GSEA and dream from raw counts

@@

# HBCC - sgACC
counts.HBCC4 = fread(synGet('syn21867866')$path, data.table=FALSE)
rownames(counts.HBCC4) = counts.HBCC4$Geneid
counts.HBCC4 = counts.HBCC4[,-c(1:6)]

@@

METADATA %>% 
  group_by(Dx) %>%  
  dplyr::summarise(count = n()) %>% 
  kable() 
```

```{r exit1, cache=FALSE}
knitr::knit_exit()
```

```{r process.counts}

# keep genes that are express in at least one combination of cohort and gender
isexpr.lst = dlply(METADATA, .(Cohort, Reported_Gender), .fun = function(mtd){
	rowSums(cpm(geneCountsMerged[,mtd$SampleID])>1) >= 0.5*ncol(geneCountsMerged[,mtd$SampleID])
	})

isexpr = do.call(rbind, isexpr.lst)
isexpr = colSums(isexpr) > 0
table(isexpr)

# normalize counts
dge = DGEList( geneCountsMerged[isexpr,] )
dge = calcNormFactors(dge) # TMM

xtabs(~ Region + Cohort, METADATA)
xtabs(~ Institution + Cohort, METADATA)

@@

# MPP DLPFC
# Dx + scale(RIN2) + (1 | Institution) + scale(ageOfDeath) + scale(RIN) + (1 | Reported_Gender) + scale(IntronicRate) + scale(IntragenicRate)

form = ~ (1|Dx) + Reported_Gender + RIN + ageOfDeath + scale(IntronicRate)*Cohort + scale(IntragenicRate)*Cohort + scale(IntergenicRate)*Cohort + scale(rRNARate)*Cohort + (1|Institution:Cohort) + (1|Region:Cohort) + (1|Individual_ID)

vobj = voomWithDreamWeights(dge, form, METADATA, save.plot=TRUE, BPPARAM = SnowParam(18))

plot_voom(vobj)
```

@@

form = ~ 0 + Dx + Reported_Gender + RIN + ageOfDeath + scale(IntronicRate)*Cohort + scale(IntragenicRate)*Cohort + scale(IntergenicRate)*Cohort + scale(rRNARate)*Cohort + (1|Institution:Cohort) + (1|Region:Cohort) + (1|Individual_ID)

# SCZ
L_SCZ = Lbase
L_SCZ['DxControl'] = -1 
L_SCZ['DxSCZ'] = 1

# BP
L_BP = Lbase
L_BP['DxControl'] = -1 
L_BP['DxBP'] = 1

# BP
L_sex = Lbase
L_sex['Reported_GenderMale'] = 1

L = cbind(	Scz = L_SCZ, 
			BP 	= L_BP, 
			Sex = L_sex)

fit_marginal = dream(vobj, form, METADATA, L=L,BPPARAM = SnowParam(18))
```

```{r plots.marginal, fig.width=10, fig.height=5}
# Disease
tab = topTable(fit_marginal, "Scz", number=Inf) %>% getGeneSymbol
fig1 = plotVolcano( tab, tab$Symbol[1:20] ) + ggtitle("SCZ")

tab = topTable(fit_marginal, "BP", number=Inf)  %>% getGeneSymbol
fig2 = plotVolcano( tab, tab$Symbol[1:20] ) + ggtitle("BP")

plot_grid( fig1, fig2, ncol=2)

# Sex  
tab = topTable(fit_marginal, "Reported_GenderMale", number=Inf) %>% getGeneSymbol
fig3 = plotVolcano( tab, tab$Symbol[1:20] ) + ggtitle("Sex")

tab2 = tab[tab$Chrom %in% 1:22,]
fig4 = plotVolcano( tab2, tab2$Symbol[1:20]) + ggtitle("Sex: autosomes")

plot_grid( fig3, fig4, ncol=2)
```

```{r gene_set_test.marginal}

library(zenith)

run_zenith = function(fit, coefs, gs.collection, n_genes_min=10, n_genes_max=5000){

	# Map from Ensembl genes in geneSets_GO to 
	# from trimmed Ensembl names from RNA-seq data 
	geneSets.index = ids2indices( gs.collection, trim_ensembl_ids(rownames(fit)))

	# filter by size of gene set
	geneSets.index = geneSets.index[sapply(geneSets.index, function(x) (length(x) >= n_genes_min) & (length(x) <= n_genes_max)) ]

	# run zenith for each coefficient
	res = lapply( coefs, function(coef){
		zenith(fit, coef, geneSets.index, squaredStats=TRUE)
		} )
	names(res) = coefs

	res
}

@@

# convert from GeneSetCollection to list used by camera and zenith
gs.GO.collection = recodeToList( gs.GO )
gs.MSigDB.collection = recodeToList( gs.MSigDB )

res_zenith_marginal.GO = run_zenith( fit_marginal, colnames(L), gs.GO.collection )
res_zenith_marginal.MSigDB = run_zenith( fit_marginal, colnames(L), gs.MSigDB.collection )
```

>>>>

@@

# # HBCC - sgACC
# counts.HBCC4 = fread(synGet('syn21867866')$path, data.table=FALSE)
# rownames(counts.HBCC4) = counts.HBCC4$Geneid
# counts.HBCC4 = counts.HBCC4[,-c(1:6)]

@@

# METADATA %>% 
#   group_by(Dx) %>%  
#   dplyr::summarise(count = n()) %>% 
#   kable() 
```

## Summarize samples
```{r summary}
METADATA %>% group_by(Dx, Institution) %>% dplyr::summarise(count = n()) %>% 
  kable(caption="Diagnosis, Institution") %>% kable_styling(full_width=FALSE)

METADATA %>% group_by( Institution, Reported_Gender) %>% dplyr::summarise(count = n()) %>% 
  kable(caption="Institution, Sex") %>% kable_styling(full_width=FALSE)

METADATA %>% group_by(Dx, Institution, Reported_Gender) %>% dplyr::summarise(count = n()) %>% 
  kable(caption="Diagnosis, Institution, Sex") %>% kable_styling(full_width=FALSE) 
```

```{r process.counts}
# keep genes that are express in at least one combination of cohort and gender
isexpr.lst = dlply(METADATA, .(Cohort, Reported_Gender), .fun = function(mtd){
	rowSums(cpm(geneCountsMerged[,mtd$SampleID])>1) >= 0.5*ncol(geneCountsMerged[,mtd$SampleID])
	})

isexpr = do.call(rbind, isexpr.lst)
isexpr = colSums(isexpr) >= 3
table(isexpr)

# Normalize each cohort separately
dge.lst = lapply( levels(METADATA$Cohort), function(Cohort){
  # normalize counts
  dge = DGEList( geneCountsMerged[isexpr,METADATA$Cohort==Cohort] )
  calcNormFactors(dge) # TMM
})
names(dge.lst) = levels(METADATA$Cohort)

# form = ~ (1|Dx) + Reported_Gender + RIN + ageOfDeath + scale(IntronicRate)*Cohort + scale(IntragenicRate)*Cohort + scale(IntergenicRate)*Cohort + scale(rRNARate)*Cohort + (1|Institution:Cohort) + (1|Region:Cohort) + (1|Individual_ID)

form = ~ Dx + Reported_Gender + RIN + ageOfDeath + scale(IntronicRate) + scale(IntragenicRate) + scale(IntergenicRate) + scale(rRNARate)

vobj.lst = lapply(dge.lst, function(dge){

  i = match(colnames(dge), rownames(METADATA))
  info = METADATA[i,]

  if( length(unique(info$Institution)) > 1 ){
    form_use = as.formula(paste('~',as.character(form)[2], "+ Institution", collapse=" "))
  }else{
    form_use = form
  }

  voomWithDreamWeights(dge, form_use, info, save.plot=TRUE, BPPARAM = SnowParam(18))
  })
names(vobj.lst) = names(dge.lst)

@@

# MPP DLPFC
# Dx + scale(RIN2) + (1 | Institution) + scale(ageOfDeath) + scale(RIN) + (1 | Reported_Gender) + scale(IntronicRate) + scale(IntragenicRate)
```

### Voom trend plots
```{r plot.voom, fig.height=5, fig.width=10}
fig1 = plot_voom(vobj.lst[[1]]) + ggtitle(names(vobj.lst)[1])
fig2 = plot_voom(vobj.lst[[2]]) + ggtitle(names(vobj.lst)[2])
plot_grid( fig1, fig2, nrow=1) 
```

@@

form = ~ Dx + Reported_Gender + RIN + ageOfDeath + scale(IntronicRate) + scale(IntragenicRate) + scale(IntergenicRate) + scale(rRNARate)

fit_marginal.lst = lapply( vobj.lst, function(vobj){

  i = match(colnames(vobj), rownames(METADATA))
  info = METADATA[i,]

  if( length(unique(info$Institution)) > 1 ){
    form_use = as.formula(paste('~',as.character(form)[2], "+ Institution", collapse=" "))
  }else{
    form_use = form
  }

  fit = dream(vobj, form, info, BPPARAM = SnowParam(18))  
  eBayes( fit )
  })
names(fit_marginal.lst) = names(vobj.lst)
```

```{r load.genesets}
library(zenith)

@@

# convert from GeneSetCollection to list used by camera and zenith
gs.GO.collection = recodeToList( gs.GO )
gs.MSigDB.collection = recodeToList( gs.MSigDB )
```

```{r run.zenith}

coefTest = c('DxSCZ', 'Reported_GenderMale')
res_zenith = lapply( names(fit_marginal.lst), function(key){

  fit = fit_marginal.lst[[key]]

  res = lapply(coefTest, function(coef){
    res1 = run_zenith( fit, coef, gs.GO.collection )
    res1$Resource = 'GO'

    res2 = run_zenith( fit, coef, gs.MSigDB.collection )
    res2$Resource = "MSigDB"

    res = rbind(res1, res2)
    res$Coef = coef
    res$Cohort = key

    res
  })
  do.call(rbind, res)
})
names(res_zenith) = names(fit_marginal.lst)
```

```{r get.signatures, fig.width=10, fig.height=5}
# Get DE signatures
df_DE_results = lapply(fit_marginal.lst, function(fit){
  res = lapply(coefTest, function(coef){
    res = topTable(fit, coef, number=Inf) 
    res$Coef = coef
    res
    })
  do.call(rbind, res) %>% getGeneSymbol
  })
names(df_DE_results) = names(fit_marginal.lst)
```

```{r test.scz, fig.height=5, fig.width=10}
# Schizophrenia testing
# extract from each chort
coef = 'DxSCZ'
figList.Scz = lapply( names(df_DE_results), function(key){
  df = df_DE_results[[key]]
  df = df[df$Coef == coef,]

  main = paste(key, coef, sep = ': ' )
  plotVolcano( df, df$Symbol[1:20] ) + ggtitle( main )
})

# volcano plots
plot_grid(plotlist=figList.Scz, nrow=1)
```

```{r meta.analysis.SCZ}
# Create list preparing for meta-analysis for SCZ
tab.lst = lapply( df_DE_results, function(df){
  df[df$Coef == coef,]
})
names(tab.lst) = names(df_DE_results)

# Run meta-analysis
df_meta_SCZ = run_meta_analysis( tab.lst, "FE") %>% getGeneSymbol
```

```{r concordance.scz, fig.height=6, fig.width=10}
# Concordance analysis
resConcord.SCZ = plot_concordance(tab.lst, showGenes = df_meta_SCZ$Symbol[1:20])

main = "Concordance of SCZ"
plot_grid(  resConcord.SCZ$fig_logFC + ggtitle(main),
            resConcord.SCZ$fig_tstat + ggtitle(main), nrow=1)

data.frame(pi1 = resConcord.SCZ$pi_single) %>% kable(caption="pi1 within dataset") %>% kable_styling(full_width=FALSE)
resConcord.SCZ$tab_pi %>% kable(caption="pi1 for replication") %>% kable_styling(full_width=FALSE)
resConcord.SCZ$tab_corr %>% kable(caption="Correlation between t-statistics") %>% kable_styling(full_width=FALSE)
```

```{r plot.meta.analysis.SCZ, fig.height=5, fig.width=10}
# volcano plot for meta-analysis
main = paste('Meta-analysis', coef, sep = ': ' )
plotVolcano( df_meta_SCZ, df_meta_SCZ$Symbol[1:20] ) + ggtitle( main )
```

```{r exit2, cache=FALSE}
knitr::knit_exit()
```

#   rmarkdown::render("combined_analysis/combine_all_DLPFC.Rmd")
